Measuring causal influence with\\ back-to-back regression

### 2019-05-23 paper/paper.tex
Changes

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage[]{neurips_2019}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F + N',\label{eq:model}
\end{equation}
%
where $N$ and $N'$ are some unobserved noise variables.
%
While the triplet of variables $X$, $N$, and $N'$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.

@@

Assuming centered noise $N$,
\begin{align*}
    \arg \min_G \left \| YG - X \right \|^2 &= \arg \min_G \left \| X - (XE + N)FG \right\|^2 \\
                                            &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
\end{align*}
%
As the noise $N$ tends to zero, the solution to the previous problem is $\hat G = (X E F)^\dagger X$, where ``$\dagger$'' denotes the Moore-Penrose pseudo-inverse.

Next, let $k$ be the number of ones in the causal influence matrix $E$.
%
Without loss of generality, assume that these are the first $k$ variables in $X$.
%
Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
%
Let $K \in \mathbb{R}^{k \times n_{x}}$ be the diagonal matrix obtained by deleting the bottom $dn_{x} - k$ lines of $E$.
%
Then, $E = K^\top K$, and $XEF = (XK^\top)(KF)$, with both $XK^\top$ and $KF$ of rank $k$.

Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalize}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%
Using this property on the minimal norm solution $\hat G$ yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
\end{aligned}
\end{equation}

Let $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$. Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{cc}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right)$, call it S.

Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.

Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.

Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$. Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).

For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$. (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$). The square norm is therefore $\lambda^2 Var(N_{active})$.

Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$. Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$. Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).

The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.

The second step is straightforward, replacing Y, we have
\begin{equation}
\begin{aligned}
\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
&= EF \hat G 
\end{aligned}
\end{equation}

Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.

>>>>

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage[]{neurips_2019}

\newtheorem{theorem}{Theorem}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F,\label{eq:model}
\end{equation}
%
where $N$ is a centered, unobserved noise variable.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.

@@

This section provides a consistency proof for B2BR in the case of low noise.

\begin{theorem}[Low-noise consistency]
    Consider the B2BR model from Equation~\ref{eq:model}.
    %
    Assume that $F$ and $X$ are full-rank on the subspace spanned by active entries in $E$. 
    %
    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2BR recovers $E$.
\end{theorem}
\begin{proof}
For centered noise $N$, the first regression satisfies:
\begin{align*}
    \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2].
\end{align*}
%
The solution to this first regression is $\hat{G} = [(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X$.

The second regression is: 
%
\begin{align*}
    \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
\end{align*}
%
The solution to this second regression is $H = EF\hat{G}$.
%
Therefore, our estimate of $E$ is:
\begin{equation}
    \hat{E} = EF[(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X.
    \label{eq:estimate}
\end{equation}
%
By the definition of the Moore-Penrose, if $(NF)^\top (NF)$ is full-rank with $\| (NF)^\top (NF) \| \to 0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
%
Next, assume that the true causal influence matrix contains $k$ ones arranged as $E = \left(\begin{array}{cc} I_k & 0 \\ 0 & 0 \end{array}\right)$, and let $E = K^\top K$, where $K\in \mathbb{R}^{k \times d_x}$.
%
    Also, let the covariance matrix of $X$ be $X^\top X = \left(\begin{array}{cc}\Sigma_{1} & \Sigma_{2} \\ \Sigma_{3} & \Sigma_{4}\end{array}\right)$, where $\Sigma_1 \in \mathbb{R}^{k\times k}$.
%
Then,
\begin{align*}
    \hat{E} &= EF(XEF)^\dagger X = K^\top K F(XK^\top K F)^\dagger X \stackrel{i)}{=} K^\top (K F) (KF)^\dagger (XK^\top)^\dagger X\\
            &\stackrel{ii)}{=} K^\top K(XK^\top)^\dagger X = E (XK^\top)^\dagger X \stackrel{iii)}{=} E (KX^\top X K^\top)^{-1} KX^\top X\\
            &= E (KX^\top X K^\top)^{-1} (\Sigma_1 \,|\, \Sigma_2)  = E \Sigma_1^{-1} (\Sigma_1 \,|\, \Sigma_2)  
            = \left(\begin{array}{cc} I_k & \Sigma_1^{-1} \Sigma_2 \\ 0 & 0 \end{array}\right),
\end{align*}
where the equalities follow due to i) a full-rank factorization of $XEF$, ii) $KF$ being full-rank yields $(KF)(KF)^\dagger = K$, and iii) $X K^\top$ being full-rank yields $(XK^\top)^\dagger = (KX^\top X K^\top)^{-1} KX^\top$.
%
Therefore, $\text{Diag}(\hat{E})$ is the desired influence causal matrix $E$. 
\end{proof}

We observe that, even in the presence of strong noise, the expression of our estimate \eqref{eq:estimate} involves the left-most multiplication with the true causal influence matrix $E$.
%
This means that our estimate will, in expectation, contain zeros for all features that do not causally influence $Y$.
%
When the variance of $N$ increases, we of course pay a price: since the norm $\| X - XEF\hat{G} \| = \| X - X\hat{H} \|$ increases, this means that the values of $\hat{E}$ associated to the causal factors will be dampened.

%%% Next, let $k$ be the number of ones in the binary diagonal causal influence matrix $E$.
%%% %
%%% That is, $\text{tr}(E) = k$.
%%% %
%%% Without loss of generality, assume that these are the first $k$ variables in $X$.
%%% %
%%% Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
%%% %
%%% Let $K \in \mathbb{R}^{k \times d_{x}}$ be the diagonal matrix obtained by deleting the bottom $d_{x} - k$ rows of $E$.
%%% %
%%% %
%%% First, we factorize $(XEF)^\dagger = ((XK^\top)(KF))^\dagger = (KF)^\dagger (XK^\top)^\dagger$.
%%% %
%%% Second, since $XK^\top$ is full rank, we write $(XK^\top)^\dagger = ((XK^\top)^\top (XK^\top))^{-1}(XK^\top)^\top$.
%%% %
%%% Then, $(XEF)^\dagger X = (KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X$.
%%% 
%%% %Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalized}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%%% %
%%% % Using this property on the minimal norm solution $\hat G$ yields
%%% % \begin{equation}
%%% % \begin{aligned}
%%% % (XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
%%% % &=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
%%% % &=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
%%% % \end{aligned}
%%% % \end{equation}
%%% 
%%% Write the covariance of $X$ as $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, with the block $\Sigma_1$ being $k\times k$. Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{c|c}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right) =: S$.
%%% 
%%% 
%%% Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
%%% 
%%% Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$. Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
%%% 
%%% For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$. (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$). The square norm is therefore $\lambda^2 Var(N_{active})$.
%%% 
%%% Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$. Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$. Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
%%% 
%%% The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} (KF)^\dagger S$.
%%% 
%%% 
%%% Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
%%% 
%%% ADD AGAIN:
%%% 
%%% Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
